Best practices for forecasting changes in biodiversity: an example using breeding birds

### 2017-08-18 manuscript/draft.Rmd
Editorial work on Methods section

In this paper, we attempt to forecast the species richness of breeding birds at
over 1,200 of sites located throughout North America, while following best
practices for ecological forecasting (Box 1). To do this, we combine 30 years of
time-series data on bird distributions from annual surveys with monthly
time-series of climate data and satellite-based remote-sensing. Datasets that
span a time scale of 30 years or more have only recently become available for
large-scale time-series based forecasting. A dataset of this size allows us to
model and assess changes a decade or more into the future in the presence of
shifts in environmental conditions on par with predicted climate change. We
compare traditional distribution modeling based approaches to spatial models of
species richness, time-series methods, and two simple baselines that predict
constant richness for each site, on average (Figure 1). All of our forecasting
models account for uncertainty and observation error, are evaluated across
different time lags using hindcasting, and are publicly archived to allow future
assessment. We discuss the implications of these practices for our understanding
of, and confidence in, the resulting forecasts, and how we can continue to build
on these approaches to improve ecological forecasting in the future.

@@

**Richness data.** Bird species richness was obtained from the North American
Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever software
[@morris2013] and rdataretriever R package [@mcglinn2017]. The BBS data was
filtered to exclude all nocturnal, cepuscular, and aquatic species [since these
species are not well sampled by BBS methods; @hurlbert2005], as well as
unidentified species, and hybrids. All data from routes that did not meet BBS
acceptable criteria were also excluded.

We used observed richness values from 1982 (the first year of complete
environmental data) to 2003 for training the models, and data from 2004 to 2013
was used for testing their performance. We only used BBS routes from the
continental United States (i.e. routes where climate data was available @prism),
and we restricted the analysis to routes that were sampled during 70% of the
years in the training period (i.e., routes with at least 16 annual
observations). The resulting dataset included `r numbers$N_runs` transect runs
of `r numbers$N_sites` unique sites, and included `r numbers$N_species` species.
Site-level richness varied from `r numbers$richness_summary$min` to `r
numbers$richness_summary$max` with an average richness of `r
numbers$richness_summary$mean` species.

**Past environmental data.** Environmental data included a combination of
elevation, bioclimatic variables and a remotely sensed vegetation index (the
normalized difference vegetation index; NDVI), all of which are known to
influence richness and distribution in the BBS data [[ref]]. For each year in
the dataset, we used the 4 km resolution PRISM data [@prism] to calculate eight
bioclimatic variables identified as relevant to bird distributions
[@harris2015]: mean diurnal range, isothermality, max temperature of the warmest
month, mean temperature of the wettest quarter, mean temperature of the driest
quarter, precipitation seasonality, precipitation of the wettest quarter, and
precipitation of the warmest quarter. Satellite-derived NDVI, a primary
correlate of richness in BBS data [@hurlbert2002], was obtained from the NDIV3g
dataset with an 8 km resolution [@pinzon2014] and was available from the time
period 1981-2013. Average summer (May, June, July) and winter (December,
January, Feburary) NDVI values were used as predictors. Elevation was from the
SRTM 90m elevation dataset [@jarvis2008] obtained using the R package raster
[@hijmans2016]. Because BBS routes are 40-km transects rather than point counts,
we used the average value of each environmental variable within a 40 km radius
of each BBS route's starting point.

**Future environmental projections.** We made long term forecasts from 2014-2050
using the CMIP5 multi-model ensemble dataset as the source for climatic
variables [@brekke2013]. Precipitation and temperature from 37 downscaled model
runs [@brekke2013, see Table S1] using the RCP6.0 scenario were averaged
together to create a single ensemble used to calculate the bioclimatic variables
for North America. For NDVI we used the per-site average values from 2000-2013
as a simple forecast. For observer effects (see below) each site was set to have
zero observer bias.

@@

For each forecasting approach, we trained two versions of the corresponding
model: one with corrections for differences among observers, and one without.
Observer effects are inherent in large data sets collected by different
observers, and are known to occur in BBS [@sauer1994]. We estimated the observer
effects (and associated uncertainty about those effects) with a linear mixed
model with observer as a random effect built in the Stan probabilistic
programming language [@stan]. Because observer and site are strongly related
(observers tend to repeatedly sample the same site), site was also included as a
random effect to ensure that inferred deviations were actually observer-related
(as opposed to being related to the sites that a given observer happened to
see). The resulting model partitions the variance in observed richness values
into site-level variance, observer-level variance, and residual variance (e.g.
variation within a site from year to year; Figure 3). The site-level estimates
can also be used directly as the "average" baseline model (see below). The
estimated observer effects can be subtracted from the richness values for a
particular observer to provide an estimate of how many species would have been
found by a "typical" observer. To incorporate uncertainty in these "corrected"
richness values into the forecasting models we collected 500 Monte Carlo samples
from the model's posterior distribution, and fit each of the downstream models
with each of the Monte Carlo samples. Each Monte Carlo sample represented a
different possible set of observer-level and site-level random effect values
across the full 30-year dataset.

![**A.** When all observers are treated the same (black points), predictions can
be poor. **B.** By accounting for systematic differences between observers
(represented by the points' colors), most models can be made more robust to
observer turnover. Note that the "naive" model is already less sensitive to
observer turnover, and does not benefit as much from modeling
it.](../figures/observer_predictions.png)

@@

We fit three sets of models that were fit to each site separately, with no
environmental information (Table 1). These models were fit to each BBS route
twice: once using the residuals from the observer model, and once using the raw
richness values. When correcting for observer effects, we averaged across 500
models that were fit separately to the 500 Monte Carlo estimates of the observer
effects, to account for our uncertainty in the true values of those effects. All
of these models use a Gaussian error distribution (rather than a count
distribution) for the reasons discussed below (see "Model evaluation").

@@

table_caption = "A summary of the six models evaluated here. The single-site
models were trained site-by-site, without access to any environmental
information. The environmental models were trained at a continental scale,
without information regarding which transects occurred at which site or during
which year. Most of the models were trained to predict richness directly.
However, the random forest SDMs were each trained to predict the occurrence of
individual species at each site, and these species-level predictions were added
up to produce richness estimates. The mistnet JSDM was trained to predict the
full species composition at each site, and the number of species in its
predictions was used as an estimate of richness."

@@

**Baseline models.** We used two simple baseline models as a basis for
comparison with the more complex models (Figure 2A). These baselines treated
site-level richness observations either as uncorrelated noise around a
site-level constant (the "average" model) or as an autoregressive model with a
single year of history (the "naive" model) [@hyndman2014]. Predictions from the
"average" model are centered on the average richness observed during training,
and the confidence intervals are narrow and constant-width. The "naive" model,
in contrast, predicts that future observations will be similar to the final
observed value (e.g. the value observed in `r timeframe$last_train_year`), and
the confidence intervals expand rapidly as the predictions extend farther into
the future. Both models' richness predictions are centered on a constant value,
so neither model can anticipate any trends in richness or any responses to
future environmental changes.

**Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
function in @forecast1] to represent a array of different time-series modeling
approaches. These models can include an autoregressive component (as in the
"naive" model, but with the possibility of longer-term dependencies in the
underlying process), a moving average component (where the noise can have serial
autocorrelation) and an integration/differencing component (so that the analysis
could be performed on sequential differences of the raw data, accommodating more
complex patterns including trends). The `auto.arima` function chooses whether to
include each of these components (and how many terms to include for each one)
using AICc [@forecast1]. Since there is no seasonal component to the BBS
time-series, we did not include a season component in these models. Otherwise we
used the default settings for this function [@forecast1].

@@

**"Macroecological" model: richness GBM.** We used a boosted regression tree
model using the `gbm` package [@ridgeway2017] to directly model species richness
as a function of environmental variables. Boosted regression trees are a form of
tree-based modeling that work by fitting thousands of small tree-structured
models sequentially, with each tree optimized to reduce the error of its
predecessors. They are flexible models that are considered well suited for
prediction [@elith2008]. This model was optimized using a Gaussian likelihood,
with a maximum interaction depth of 5, shrinkage of 0.015, and up to 10,000
trees. The number of trees used for prediction was selected using the "out of
bag" estimator; this number averaged `r numbers$n_gbm[1]` for the non-observer
data and `r numbers$n_gbm[2]` for the observer-corrected data. [[add comma
separators for thousands place]]

@@

**Joint Species Distribution Model: mistnet.** Joint species distribution models
(JSDMs) are a new approach that makes predictions about the full composition of
a community instead of modeling each species independently as above. JSDMs
remove the assumed independence among species and explicitly account for the
possibility that a site will be much more (or less) suitable for birds in
general (or particular groups of birds) than one would expect based on the
available environmental measurements alone. As a result, JSDMs do a better job
of representing our uncertainty about richness, while stacked SDMs underestimate
it [@harris2015; [[other refs, e.g., Clark]]]. We used the `mistnet` package
[@harris2015] because it is the only JSDM that describes species' environmental
associations with nonlinear functions.

@@

We defined model performance for all models in terms of continuous Gaussian
errors, instead of using discrete count distributions. Variance in species
richness within sites was lower than predicted by several common count models,
such as the Poisson or binomial (i.e. richness was underdispersed for individual
sites), so these count models would have had difficulty fitting the data [cf.
@calabrese2014]. The use of a continuous distribution is adequate here, since
richness had a relatively large mean (`r numbers$richness_summary$mean`) and all
models already produce continuous richness estimates (due to summing SDM or JSDM
probabilities). When a model was run multiple times for the purpose of
correcting for observer effects, we used the mean of those runs' point estimates
as our final point estimate and we calculated the uncertainty using the law of
total variance (i.e. the average of the model runs' variance, plus the variance
in the point estimates).

We evaluated each model's forecasts using the data for each year between `r
timeframe$last_train_year` and `r timeframe$end_yr`. We used three metrics for
evaluating performance: 1) root-mean-square error (RMSE) to determine how far,
on average, the models' predictions were from the observed value; 2) the 95%
prediction interval coverage to determine how well the models predicted the
range of possible outcomes; and 3) deviance (i.e. negative 2 times the Gaussian
log-likelihood) as an integrative measure of fit incorporating good point
estimates, precision, and coverage.  In addition to evaluating forecast
performance in general, we evaluated how performance changed as the time horizon
of forecasting increased by plotting performance metrics against year. Finally,
we decomposed each model's squared error into two components: one described the
squared error associated with site-level means, while the other described the
squared error associated with annual fluctuations in richness within a site.
This decomposition tells us the extent to which each model's error depends on
consistent differences among sites versus changes in site-level richness from
year to year.

>>>>

In this paper, we attempt to forecast the species richness of breeding birds at
over 1,200 of sites located throughout North America, while following best
practices for ecological forecasting (Box 1). To do this, we combine 32 years of
time-series data on bird distributions from annual surveys with monthly
time-series of climate data and satellite-based remote-sensing. Datasets that
span a time scale of 30 years or more have only recently become available for
large-scale time-series based forecasting. A dataset of this size allows us to
model and assess changes a decade or more into the future in the presence of
shifts in environmental conditions on par with predicted climate change. We
compare traditional distribution modeling based approaches to spatial models of
species richness, time-series methods, and two simple baselines that predict
constant richness for each site, on average (Figure 1). All of our forecasting
models account for uncertainty and observation error, are evaluated across
different time lags using hindcasting, and are publicly archived to allow future
assessment. We discuss the implications of these practices for our understanding
of, and confidence in, the resulting forecasts, and how we can continue to build
on these approaches to improve ecological forecasting in the future.

@@

**Richness data.** Bird species richness was obtained from the North American
Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever Python
package [@morris2013] and rdataretriever R package [@mcglinn2017]. The BBS data
was filtered to exclude all nocturnal, cepuscular, and aquatic species [since
these species are not well sampled by BBS methods; @hurlbert2005], as well as
unidentified species, and hybrids. All data from surveys that did not meet BBS
quality criteria were also excluded.

We used observed richness values from 1982 (the first year of complete
environmental data) to 2003 to train the models, and from 2004 to 2013 to test
their performance. We only used BBS routes from the continental United States
(i.e. routes where climate data was available @prism), and we restricted the
analysis to routes that were sampled during 70% of the years in the training
period (i.e., routes with at least 16 annual observations). The resulting
dataset included `r numbers$N_runs` annual surveys of `r numbers$N_sites` unique
sites, and included `r numbers$N_species` species.  Site-level richness varied
from `r numbers$richness_summary$min` to `r numbers$richness_summary$max` with
an average richness of `r numbers$richness_summary$mean` species.

**Past environmental data.** Environmental data included a combination of
elevation, bioclimatic variables and a remotely sensed vegetation index (the
normalized difference vegetation index; NDVI), all of which are known to
influence richness and distribution in the BBS data [[ref]]. For each year in
the dataset, we used the 4 km resolution PRISM data [@prism] to calculate eight
bioclimatic variables identified as relevant to bird distributions
[@harris2015]: mean diurnal range, isothermality, max temperature of the warmest
month, mean temperature of the wettest quarter, mean temperature of the driest
quarter, precipitation seasonality, precipitation of the wettest quarter, and
precipitation of the warmest quarter. Satellite-derived NDVI, a primary
correlate of richness in BBS data [@hurlbert2002], was obtained from the NDIV3g
dataset with an 8 km resolution [@pinzon2014] and was available from 1981-2013.
Average summer (May, June, July) and winter (December, January, Feburary) NDVI
values were used as predictors. Elevation was from the SRTM 90m elevation
dataset [@jarvis2008] obtained using the R package raster
[@hijmans2016]. Because BBS routes are 40-km transects rather than point counts,
we used the average value of each environmental variable within a 40 km radius
of each BBS route's starting point.

**Future environmental projections.** We made long term forecasts from 2014-2050
using the CMIP5 multi-model ensemble dataset as the source for climate
variables [@brekke2013]. Precipitation and temperature from 37 downscaled model
runs [@brekke2013, see Table S1] using the RCP6.0 scenario were averaged
together to create a single ensemble used to calculate the bioclimatic variables
for North America. For NDVI we used the per-site average values from 2000-2013
as a simple forecast. For observer effects (see below) each site was set to have
zero observer bias.

@@

Observer effects are inherent in large data sets collected by different
observers, and are known to occur in BBS [@sauer1994]. For each forecasting
approach, we trained two versions of the corresponding model: one with
corrections for differences among observers, and one without (Figure 2). We
estimated the observer effects (and associated uncertainty about those effects)
with a linear mixed model, with observer as a random effect, built in the Stan
probabilistic programming language [@stan]. Because observer and site are
strongly related (observers tend to repeatedly sample the same site), site was
also included as a random effect to ensure that inferred deviations were
actually observer-related (as opposed to being related to the sites that a given
observer happened to see). The resulting model partitions the variance in
observed richness values into site-level variance, observer-level variance, and
residual variance (e.g.  variation within a site from year to year). The
site-level estimates can also be used directly as the "average" baseline model
(see below). The estimated observer effects can be subtracted from the richness
values for a particular observer to provide an estimate of how many species
would have been found by a "typical" observer. To incorporate uncertainty in
these "corrected" richness values into the forecasting models we collected 500
Monte Carlo samples from the model's posterior distribution, and fit each of the
downstream models with each of the Monte Carlo samples. Each Monte Carlo sample
represented a different possible set of observer-level and site-level random
effect values across the full 32-year dataset.

![**A.** Model predictions for [[STATE]] route [[ROUTE NUMBER]] when all
observers are treated the same (black points). **B.** Model predictions for the
same route when accounting for systematic differences between observers
(represented by the points' colors). In this example most models are made more
robust to observer turnover. Note that the "naive" model is less sensitive to
observer turnover, and does not benefit as much from modeling
it.](../figures/observer_predictions.png)

@@

Three of models used in this study were fit to each site separately, with no
environmental information (Table 1). These models were fit to each BBS route
twice: once using the residuals from the observer model, and once using the raw
richness values. When correcting for observer effects, we averaged across 500
models that were fit separately to the 500 Monte Carlo estimates of the observer
effects, to account for our uncertainty in the true values of those effects. All
of these models use a Gaussian error distribution (rather than a count
distribution) for reasons discussed below (see "Model evaluation").

@@

table_caption = "Six forecasting models. Single-site models were trained
site-by-site, without environmental data. Environmental models were trained
using all sites together, without information regarding which transects occurred
at which site or during which year. Most of the models were trained to predict
richness directly. This mirrors the standard application of these
techniques. Separate random forest SDMs were fit for each species and used to
predict the probability of that species occurring at each site. The
species-level probabilities at a site were summed to predict richness. The
mistnet JSDM was trained to predict the full species composition at each site,
and the number of species in its predictions was used as an estimate of
richness."

@@

**Baseline models.** We used two simple baseline models as a basis for
comparison with the more complex models (Figure 2A). These baselines treated
site-level richness observations either as uncorrelated noise around a
site-level constant (the "average" model) or as an autoregressive model with a
single year of history [the "naive" model, @hyndman2014]. Predictions from the
"average" model are centered on the average richness observed during training,
and the confidence intervals are narrow and constant-width. The "naive" model,
in contrast, predicts that future observations will be similar to the final
observed value (e.g., in our hindcasts the value observed in `r
timeframe$last_train_year`), and the confidence intervals expand rapidly as the
predictions extend farther into the future. Both models' richness predictions
are centered on a constant value, so neither model can anticipate any trends in
richness or any responses to future environmental changes.

**Time series models.** We used Auto-ARIMA models [based on the `auto.arima`
function in @forecast1] to represent an array of different time-series modeling
approaches. These models can include an autoregressive component (as in the
"naive" model, but with the possibility of longer-term dependencies in the
underlying process), a moving average component (where the noise can have serial
autocorrelation) and an integration/differencing component (so that the analysis
could be performed on sequential differences of the raw data, accommodating more
complex patterns including trends). The `auto.arima` function chooses whether to
include each of these components (and how many terms to include for each one)
using AICc [@forecast1]. Since there is no seasonal component to the BBS
time-series, we did not include a season component in these models. Otherwise we
used the default settings for this function [@forecast1].

@@

**"Macroecological" model: richness GBM.** We used a boosted regression tree
model using the `gbm` package [@ridgeway2017] to directly model species richness
as a function of environmental variables. Boosted regression trees are a form of
tree-based modeling that work by fitting thousands of small tree-structured
models sequentially, with each tree optimized to reduce the error of its
predecessors. They are flexible models that are considered well suited for
prediction [@elith2008]. This model was optimized using a Gaussian likelihood,
with a maximum interaction depth of 5, shrinkage of 0.015, and up to 10,000
trees. The number of trees used for prediction was selected using the "out of
bag" estimator; this number averaged `r numbers$n_gbm[1]` for the non-observer
data and `r numbers$n_gbm[2]` for the observer-corrected data.

@@

**Joint Species Distribution Model: mistnet.** Joint species distribution models
(JSDMs) are a new approach that makes predictions about the full composition of
a community instead of modeling each species independently as above
[@warton2015]. JSDMs remove the assumed independence among species and
explicitly account for the possibility that a site will be much more (or less)
suitable for birds in general (or particular groups of birds) than one would
expect based on the available environmental measurements alone. As a result,
JSDMs do a better job of representing uncertainty about richness than stacked
SDMs [@harris2015; [[other refs, e.g., Clark]]]. We used the `mistnet` package
[@harris2015] because it is the only JSDM that describes species' environmental
associations with nonlinear functions.

@@

We defined model performance for all models in terms of continuous Gaussian
errors, instead of using discrete count distributions. Variance in species
richness within sites was lower than predicted by several common count models,
such as the Poisson or binomial (i.e. richness was underdispersed for individual
sites), so these count models would have had difficulty fitting the data [cf.
@calabrese2014]. The use of a continuous distribution is adequate here, since
richness had a relatively large mean (`r numbers$richness_summary$mean`) and all
models produce continuous richness estimates. When a model was run multiple
times for the purpose of correcting for observer effects, we used the mean of
those runs' point estimates as our final point estimate and we calculated the
uncertainty using the law of total variance (i.e. the average of the model runs'
variance, plus the variance in the point estimates).

We evaluated each model's forecasts using the data for each year between `r
timeframe$last_train_year + 1` and `r timeframe$end_yr`. We used three metrics
for evaluating performance: 1) root-mean-square error (RMSE) to determine how
far, on average, the models' predictions were from the observed value; 2) the
95% prediction interval coverage to determine how well the models predicted the
range of possible outcomes; and 3) deviance (i.e. negative 2 times the Gaussian
log-likelihood) as an integrative measure of fit incorporating good point
estimates, precision, and coverage. In addition to evaluating forecast
performance in general, we evaluated how performance changed as the time horizon
of forecasting increased by plotting performance metrics against year. Finally,
we decomposed each model's squared error into two components: the squared error
associated with site-level means and the squared error associated with annual
fluctuations in richness within a site. This decomposition describes the extent
to which each model's error depends on consistent differences among sites versus
changes in site-level richness from year to year.
